---
title: "Final NBA Project"
author: "Marc Fridson"
---

Reload scraped data from CSVs and Union NBA Team Rosters 2001-2024 with 1979-2000

```{r}
library(rvest)
library(dplyr)
library(purrr)
library(stringr)
library(readr)
library(dplyr)
library(tibble)

all_nba_rosters <- read.csv("all_nba_rosters.csv")
historical_rosters <- read.csv("historical_rosters.csv")
player_stats_tbl <- read.csv("player_stats.csv")
team_stats_tbl <- read.csv("team_stats.csv")
nba_champions <- read.csv("nba_champions.csv")


all_nba_rosters <- all_nba_rosters %>%
  union_all(historical_rosters)

draft_team <- all_nba_rosters %>%
  filter(Exp == "R")
```

Combine draft, roster and championship information

```{r}

team_mapping <- data.frame(
  abbreviation = c("ATL", "BOS", "BKN", "CHA", "CHI", "CLE", "DAL", "DEN", "DET", "GSW",
                   "HOU", "IND", "LAC", "LAL", "MEM", "MIA", "MIL", "MIN", "NOP", "NYK",
                   "OKC", "ORL", "PHI", "PHX", "POR", "SAC", "SAS", "TOR", "UTA", "WAS"),
  full_name = c("Atlanta Hawks", "Boston Celtics", "Brooklyn Nets", "Charlotte Hornets", "Chicago Bulls", "Cleveland Cavaliers", 
                "Dallas Mavericks", "Denver Nuggets", "Detroit Pistons", "Golden State Warriors",
                "Houston Rockets", "Indiana Pacers", "Los Angeles Clippers", "Los Angeles Lakers", "Memphis Grizzlies", "Miami Heat", 
                "Milwaukee Bucks", "Minnesota Timberwolves", "New Orleans Pelicans", "New York Knicks",
                "Oklahoma City Thunder", "Orlando Magic", "Philadelphia 76ers", "Phoenix Suns", "Portland Trail Blazers", 
                "Sacramento Kings", "San Antonio Spurs", "Toronto Raptors", "Utah Jazz", "Washington Wizards")
)

all_nba_rosters <- all_nba_rosters %>%
  filter(Season >= 2001)%>%
  left_join(team_mapping, by = c("Team" = "abbreviation"))

nba_data <- all_nba_rosters %>%
  mutate(is_on_draft_team = if_else(
    mapply(function(Player, Birth.Date, Team) any(
      draft_team$Player == Player & draft_team$Birth.Date == Birth.Date & draft_team$Team == Team
    ), Player, Birth.Date, Team),
    TRUE, FALSE
  ))

nba_data <- nba_data %>%
  mutate(is_champion = if_else(
    mapply(function(full_name, Season) any(nba_champions$champion == full_name & nba_champions$season == Season),full_name, Season),TRUE, FALSE))

write_csv(nba_data, "nba_data.csv")

nba_data
```

### Research question

Do NBA teams that win a championship have a statistically higher or lower percentage of regular-season points scored by players they originally drafted compared to teams that do not win championships?

Additionally, when a new Collective Bargaining Agreement(CBA) has been signed in the NBA, has this significantly shifted the proportion of players on the roster drafted by that team and proportion of points scored on championship teams by players drafted by that team in that season. The intention of each CBA has been to create league parity and for players to be incentivized to stay with the teams that drafted them.

Have the CBAs cumulatively had an effect on the percentage of points scored by drafted players on championship teams? Which CBA had the most significant effect on the percentage of points scored by drafted players on championship teams?

Additionally, I will analyze total plus minus (Sum of how many points a team wins each game by (+) or loses by (-) of the NBA Champion that season to see if there is statistical significance compared to the following groups: all teams, top 16 highest plus minus values for the season, top 8 highest plus minus and top 4 highest plus minus.

### Cases

**What are the cases, and how many are there?**

We have 24 nba champions (total cases). Those 24 champions are divided into 4 time periods (2001-2005, 2005-2011, 2011-2016, 2016-2023).

### Data collection

**Describe the method of data collection.**

Roster by Year and Champions by Year ([https://www.basketball-reference.com](https://www.basketball-reference.com/teams/ATL/players.html){.uri})

Player Bios, Player Stats, Team Stats (<https://www.nba.com/stats/>)

### Type of study

**What type of study is this (observational/experiment)?**

This is an observational study.

### Data Source

-   **Basketball-Reference.com**: <https://www.basketball-reference.com/>

-   **NBA Official Website**: <https://www.nba.com/stats/>

### Response

**What is the response variable, and what type is it (numerical/categorical)?**

The response variable is the **percentage of total team points scored by homegrown players** during the regular season. It is **numerical**.

### Explanatory

**What is the explanatory variable, and what type is it (numerical/categorical)?**

The explanatory variable is whether the team **won the championship** in that season. It is **categorical** (1 = Champion, 0 = Non-Champion).

### Relevant summary statistics

Aggregate data at the team level

```{r}
nba_data_grouped <- nba_data %>%
  group_by(full_name, Season, is_champion) %>%
  summarize(
    player_count = n(),
    draft_team_count = sum(is_on_draft_team == TRUE),
    pct_homegrown = draft_team_count / player_count
  )

nba_data_grouped
```

Perform Hypothesis with 90% Confidence Level

```{r}

champion_data <- nba_data_grouped %>% filter(is_champion == TRUE) %>% pull(pct_homegrown)
non_champion_data <- nba_data_grouped %>% filter(is_champion == FALSE) %>% pull(pct_homegrown)

t_test_result <- t.test(champion_data, non_champion_data, conf.level = 0.90)

t_test_result
```

Simply looking at the % of drafted players on championship teams versus non-championship teams is not significantly different. The final project will focus more on the % of points scored by drafted players to essentially use points as a weighted average for the weighted average drafted players contribute to their teams.

```{r}

player_points_data <- player_stats_tbl %>%
  group_by(namePlayer, nameTeam, yearSeason) %>%
  summarize(total_pts = sum(pts, na.rm=TRUE), .groups = "drop")

joined_data <- player_points_data %>%
  inner_join(nba_data, by = c("namePlayer" = "Player", 
                              "nameTeam" = "full_name", 
                              "yearSeason" = "Season"))

team_points <- joined_data %>%
  group_by(nameTeam, yearSeason, is_champion) %>%
  summarize(
    team_total_pts = sum(total_pts, na.rm=TRUE),
    drafted_total_pts = sum(ifelse(is_on_draft_team == TRUE, total_pts, 0), na.rm=TRUE),
    pct_homegrown_points = drafted_total_pts / team_total_pts,
    drafted_player_count = sum(is_on_draft_team == TRUE),
    player_count = n(),
    .groups = "drop"
)

team_points <- team_points %>%
  mutate(cba_era = case_when(
    yearSeason <= 2005 ~ "pre_2005_CBA",
    yearSeason >= 2006 & yearSeason <= 2011 ~ "CBA_2006_2011",
    yearSeason >= 2012 & yearSeason <= 2016 ~ "CBA_2012_2016",
    yearSeason >= 2017 ~ "CBA_2017_onward"
  ))
```

```{r}
champion_pct <- team_points %>% filter(is_champion == TRUE) %>% pull(pct_homegrown_points)
non_champion_pct <- team_points %>% filter(is_champion == FALSE) %>% pull(pct_homegrown_points)

t_test_result <- t.test(champion_pct, non_champion_pct, conf.level = 0.80)
t_test_result
```

```{r}
champion_data_cba <- team_points %>% filter(is_champion == TRUE, !is.na(cba_era))

anova_model <- aov(pct_homegrown_points ~ cba_era, data=champion_data_cba)
summary(anova_model)
```

```{r}
TukeyHSD(anova_model)
```

The most significant drop occurred between the CBA that ended in 2005 and the CBA in effect from 2012-2016.

```{r}
library(ggplot2)
ggplot(champion_data_cba, aes(x=cba_era, y=pct_homegrown_points)) +
  geom_boxplot() +
  labs(title="Homegrown Points Percentage by CBA Era (Champions)",
       x="CBA Era",
       y="Percentage of Points from Drafted Players") + theme_minimal()
```

```{r}
champion_key <- nba_champions %>%
  rename(nameTeam = champion, yearSeason = season) %>%
  distinct(nameTeam, yearSeason) %>%
  mutate(champ_key = paste(nameTeam, yearSeason, sep = "_")) %>%
  pull(champ_key)

team_ppg <- team_stats_tbl %>%
  group_by(nameTeam, yearSeason) %>%
  summarize(avg_ppg = mean(ptsTeam, na.rm=TRUE), .groups = "drop") %>%

  mutate(champ_key = paste(nameTeam, yearSeason, sep = "_")) %>%

  mutate(is_champion = if_else(champ_key %in% champion_key, TRUE, FALSE)) %>%
  select(nameTeam, yearSeason, avg_ppg, is_champion)


champion_team_ppg <- team_ppg %>%
  filter(is_champion == TRUE)

champion_ppg <- champion_team_ppg$avg_ppg
length(champion_ppg) 

non_champion_ppg <- team_ppg %>%
  filter(is_champion == FALSE) %>%
  pull(avg_ppg)

t_test_ppg <- t.test(champion_ppg, non_champion_ppg, conf.level = 0.90)
t_test_ppg
```

**Plus/Minus Analysis**

```{r}

champion_key <- nba_champions %>%
  rename(nameTeam = champion, yearSeason = season) %>%
  distinct(nameTeam, yearSeason) %>%
  mutate(champ_key = paste(nameTeam, yearSeason, sep = "_")) %>%
  pull(champ_key)


team_plusminus <- team_stats_tbl %>%
  group_by(nameTeam, yearSeason) %>%
  summarize(total_plusminus = sum(plusminusTeam, na.rm=TRUE), .groups = "drop") %>%

  mutate(champ_key = paste(nameTeam, yearSeason, sep = "_"),
         is_champion = if_else(champ_key %in% champion_key, TRUE, FALSE)) %>%
  select(nameTeam, yearSeason, total_plusminus, is_champion)
```

```{r}
champion_plusminus <- team_plusminus %>%
  filter(is_champion == TRUE) %>%
  select(yearSeason, total_plusminus) %>%
  rename(champion_plusminus = total_plusminus)

top_plusminus <- team_plusminus
```

```{r}
top16_plusminus <- top_plusminus %>%
  group_by(yearSeason) %>%
  arrange(desc(total_plusminus)) %>%
  slice_head(n=16) %>%
  ungroup() %>%
  pull(total_plusminus)
```

```{r}
top8_plusminus <- top_plusminus %>%
  group_by(yearSeason) %>%
  arrange(desc(total_plusminus)) %>%
  slice_head(n=8) %>%
  ungroup() %>%
  pull(total_plusminus)
```

```{r}
top4_plusminus <- top_plusminus %>%
  group_by(yearSeason) %>%
  arrange(desc(total_plusminus)) %>%
  slice_head(n=4) %>%
  ungroup() %>%
  pull(total_plusminus)
```

```{r}
champion_values <- champion_plusminus$champion_plusminus

t_test_top16 <- t.test(champion_values, top16_plusminus, conf.level=0.90)
t_test_top16
```

```{r}
t_test_top8 <- t.test(champion_values, top8_plusminus, conf.level=0.90)
t_test_top8
```

```{r}
t_test_top4 <- t.test(champion_values, top4_plusminus, conf.level=0.90)
t_test_top4
```

```{r}
# Aggregate at team-season level
team_points <- joined_data %>%
  group_by(nameTeam, yearSeason, is_champion) %>%
  summarize(
    team_total_pts = sum(total_pts, na.rm=TRUE),
    drafted_total_pts = sum(ifelse(is_on_draft_team == TRUE, total_pts, 0), na.rm=TRUE),
    pct_homegrown_points = drafted_total_pts / team_total_pts,
    drafted_player_count = sum(is_on_draft_team == TRUE),
    player_count = n(),
    .groups = "drop"
)

# Define CBA Eras
team_points <- team_points %>%
  mutate(cba_era = case_when(
    yearSeason <= 2005 ~ "pre_2005_CBA",
    yearSeason >= 2006 & yearSeason <= 2011 ~ "CBA_2006_2011",
    yearSeason >= 2012 & yearSeason <= 2016 ~ "CBA_2012_2016",
    yearSeason >= 2017 ~ "CBA_2017_onward"
  ))
```

```{r}

team_stats_tbl <- team_stats_tbl %>%
  mutate(ptsOpponent = ptsTeam - plusminusTeam)

team_wins <- team_stats_tbl %>%
  group_by(nameTeam, yearSeason) %>%
  summarize(
    wins = sum(outcomeGame == "W"),
    losses = sum(outcomeGame == "L"),
    win_percentage = wins / (wins + losses),
    avg_points_for = mean(ptsTeam, na.rm=TRUE),
    avg_points_against = mean(ptsOpponent, na.rm=TRUE),
    .groups = "drop"
)

model_data <- team_points %>%
  inner_join(team_wins, by = c("nameTeam", "yearSeason"))

model_data <- model_data %>%
  mutate(
    cba_era = factor(cba_era),
    is_champion = factor(is_champion, levels = c(FALSE, TRUE), labels = c("0","1"))
  )

model_data$is_champion <- as.factor(model_data$is_champion)
levels(model_data$is_champion) <- c("NoChampion", "Champion")
```

```{r}
class_counts <- table(model_data$is_champion)
total_counts <- sum(class_counts)
class_weights <- total_counts / (length(class_counts) * class_counts)

model_data <- model_data %>%
  mutate(weights = ifelse(is_champion == "Champion",
                          class_weights["Champion"],
                          class_weights["NoChampion"]))
```

```{r}
library(caret)
set.seed(123)
train_index <- createDataPartition(model_data$is_champion, p = 0.8, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]
```

```{r}
train_data <- na.omit(train_data)
```

```{r}
f1 <- function(data, lev = NULL, model = NULL) {
  precision <- posPredValue(data$pred, data$obs, positive = lev[1])
  recall <- sensitivity(data$pred, data$obs, positive = lev[1])
  f1_val <- 2 * (precision * recall) / (precision + recall)
  names(f1_val) <- "F1"
  return(f1_val)
}
```

```{r}
objControl <- trainControl(
  method = "cv",                   
  number = 5,                     
  summaryFunction = f1,           
  classProbs = TRUE,               
  savePredictions = TRUE
)
```

```{r}
set.seed(123)
caretLogitModel <- train(
  is_champion ~ cba_era + drafted_player_count + pct_homegrown_points + win_percentage + avg_points_for + avg_points_against,
  data = train_data,
  method = "glm",
  family = binomial,
  weights = train_data$weights,
  trControl = objControl,
  metric = "F1"                 
)
```

```{r}
test_probs <- predict(caretLogitModel, newdata = test_data, type = "prob")[, "Champion"]
test_pred <- predict(caretLogitModel, newdata = test_data)
confusionMatrix(test_pred, test_data$is_champion, positive = "Champion")
```

```{r}
library(pROC)
roc_obj <- roc(test_data$is_champion, test_probs, levels = c("NoChampion", "Champion"), direction = "<")
```

```{r}
ggroc(roc_obj) +
  ggtitle(paste("ROC Curve (AUC =", round(auc(roc_obj), 2), ")")) +
  theme_minimal()
```

```{r}
library(yardstick)

# Load necessary libraries
library(yardstick)
library(ggplot2)

# Assuming 'pr_data' is your data frame with columns 'truth' and 'prob'
# Ensure 'truth' is a factor with the positive class as the first level
pr_data <- pr_data %>%
  mutate(truth = factor(truth, levels = c("Champion", "NoChampion")))

# Compute the precision-recall curve
pr_curve_data <- pr_curve(pr_data, truth = truth, prob, event_level = "first")

# Plot the precision-recall curve
autoplot(pr_curve_data) +
  ggtitle("Precision-Recall Curve") +
  theme_minimal()
```
